
\documentclass[11pt, a4paper]{awesome-cv}
\geometry{left=2cm, top=1.2cm, right=2cm, bottom=2cm, footskip=.5cm}
\fontdir[fonts/]

\newcommand*{\sectiondir}{resume/}

\name{David H.}{Silver}
\position{Researcher {\enskip\cdotp\enskip} Computational Biologist {\enskip\cdotp\enskip} AI/ML Expert {\enskip\cdotp\enskip} Academic Collaborator}
\address{NY, USA}

\mobile{+1-929-782-2060}
\email{silverdavi@gmail.com}
\linkedin{davidsilveril}
% \github{username} % Uncomment if you have GitHub

\begin{document}

\makecvheader

\makecvfooter
  {\today}
  {David H. Silver~~~·~~~Curriculum Vitae}
  {\thepage}

\cvsection{Professional Summary}
\begin{cvparagraph}
An interdisciplinary researcher bridging computational biology, geometric machine learning, and neuroscience with practical applications in healthcare and technology. With 12+ peer-reviewed publications in journals including \textit{Nature}, \textit{PNAS}, and \textit{Molecular BioSystems}, and 15+ patents in AI and medical imaging, I combine theoretical research with industry implementation.

My academic foundation at Technion---Israel Institute of Technology encompasses computational evolution, neurocognitive research, and geometric computer vision. I have extensive experience mentoring researchers and collaborating with leading academics at Oxford, Vienna, NYU, and Technion, with particular interest in remote academic collaboration and interdisciplinary research.
\end{cvparagraph}

\cvsection{Professional Experience}
\cventry
    {Head of AI/ML - \href{https://www.rheafertility.com/}{Rhea Labs}}
    {2023 - Present}
    {
      Leading AI/ML initiatives at Rhea Labs, a company focused on elevating patients' reproductive health journeys through innovative technology. Through the acquisition and integration of Embryonics (where I served as CEO/CTO), I oversee the development of AI-powered diagnostic tools and patient journey management technologies.
    }

\cventry
    {Head of AI/ML - \href{https://www.canotera.com}{Canotera}}
    {July 2023 - Present}
    {
      Canotera is revolutionizing legal tech by developing a legal dispute outcomes predictor to replace costly, time-consuming, and biased litigation with a tech-driven system for quick and fair resolutions. As Head of AI/ML, I lead the AI initiatives and collaborate with a talented team to build innovative solutions that open significant opportunities in litigation funding, insurance, and law firms.
    }


\cventry
    {Chief Technology Officer (CTO) - \href{https://www.aka-food.com/}{AKA FOODS}}
    {July 2021 - December 2023}
    {
      AKA FOODS focuses on reimagining plant-based food development from farm to fork, using deep learning to create foods with authentic taste and texture. At AKA FOODS, I initiated the company's deep-learning strategies and R\&D roadmap and led the development of a generative AI system for recipe generation, seamlessly integrating it with our food lab.
    }

\cventry
    {Chief Executive Officer (CEO) \& CTO - \href{https://embryonics.me}{Embryonics}}
    {December 2019 - Present}
    {
      Embryonics is a pioneering fertility-tech company aiming to improve IVF success rates with Geometric ML methodologies. My main role as CTO was developing and executing the company’s strategy and bridging academic research with our algorithm group to transform fertility clinics' procedures. As CEO I have led to a successful acquisition by Rhea in 2023.
    }

\cventry
    {ML Researcher - Apple Inc.}
    {August 2016 - December 2019}
    {
      At Apple, I engaged in classical signal processing algorithms and machine learning research in both software and hardware projects, contributing to cross-disciplinary projects in various sensors and technologies.
    }

\cventry
    {Algorithm Engineer - Intel Corporation}
    {July 2013 - August 2016}
    {
      At Intel I was a computer vision algorithm engineer in the RealSense depth sensing group, in which I developed solutions from the theoretical R\&D phase to ASIC-compatible implementations.
    }

\cventry
    {Research Assistant - Computational Evolution \& Development lab, Technion}
    {2010-2013}
    {
      In the Computational Evolution \& Development lab at Technion, I focused on algorithms in computational biology and high-throughput data analysis, contributing to genomics and gene-expression research under Prof. Itai Yanai.
    }

\cventry
    {Research Assistant - Evoked Potentials Laboratory, Technion}
    {2008-2010}
    {
      Conducted research in Prof. Hillel Pratt's Evoked Potentials Laboratory at Technion---Israel Institute of Technology. Prof. Pratt, who received his PhD from Hebrew University and completed a post-doc at UC Irvine, directs research on neural signal processing and cognitive electrophysiology. Contributed to studies on evoked potentials, neural signal analysis, and brain-computer interface development using EEG technologies.
    }

\cventry
    {Research Assistant - VR and Neurocognition lab, Technion}
    {2009-2010}
    {
      At the VR and Neurocognition lab, Technion, I assisted Prof. Miriam Reiner in constructing VR scenes and analyzing brain data, contributing to the intersection of virtual reality and neurocognitive studies.
    }

\cventry
    {Research Assistant - Intelligent Systems lab, CS Dept., Technion}
    {2007-2009}
    {
      Collaborating with Prof. Michael Lindenbaum at the Intelligent Systems lab, I applied classical machine learning techniques in image processing.
    }

\newpage
\cvsection{Education}
\cventry
    {Microsoft Research PhD Fellowship, Cambridge, UK}
    {2011--2013}
    {Awarded a prestigious fellowship supporting doctoral research in Biomathematics, Bioinformatics, and Computational Biology, in collaboration with the Technion Graduate School and Microsoft Research Cambridge, UK. Focused on high-throughput data analysis, evolutionary modeling, and the mathematical foundations of biological systems.}


\cventry
    {B.Sc. toward Mathematics \& Biology, Unfinished}
    {2007--2013}
    {Accepted into the Rothschild Scholars Program for Excellence at the Technion---Israel Institute of Technology. Engaged in interdisciplinary coursework and research spanning mathematics, biology, and computer science, with early research experience in computational evolution, neurocognition, and machine learning. Transitioned to applied research and industry roles prior to degree completion.}

\cvsection{Academic Interests and Teaching}

\begin{cvparagraph}
\textbf{Research Focus Areas:}
• Geometric Machine Learning and Computational Geometry
• AI Applications in Healthcare and Biology
• Neuroscience and Brain-Computer Interfaces
• Interdisciplinary Research Methodology
• Technology Transfer and Commercialization

\textbf{Teaching and Mentorship Experience:}
• Supervised graduate and undergraduate research projects at Technion
• Mentored junior researchers in AI/ML applications at Embryonics, AKA Foods, and Rhea Labs
• Developed educational content for interdisciplinary research methods
• Experience in research supervision and academic collaboration
• Open to remote mentoring and student guidance in AI and computational neuroscience
\end{cvparagraph}

\cvsection{Publications}
\begin{cvpublications}
  \cvpublication
  {A Novel Lineage of Myoviruses Infecting Cyanobacteria is Widespread in the Oceans}
  {PNAS, Jan 23, 2012 (Co-author)}

  \cvpublication
  {Identifying Functional Links Between Genes by Evolutionary Transcriptomics}
  {Molecular BioSystems, Jun 13, 2012 (First author)}

  \cvpublication
  {ELOPER: Elongation of Paired-End Reads as a Pre-Processing Tool for Improved de Novo Genome Assembly}
  {Bioinformatics, Apr 19, 2013 (First author)}

  \cvpublication
  {The Mid-Developmental Transition and the Evolution of Animal Body Plans}
  {Nature, Feb 17, 2016 (Co-author)}

  \cvpublication
  {The Gene Regulatory Program of Acrobeloides nanus Reveals Conservation of Phylum-Specific Expression}
  {PNAS, Apr 6, 2018 (Co-author)}

  \cvpublication
  {Intel® RealSense™ SR300 Coded Light Depth Camera}
  {IEEE Transactions on Pattern Analysis and Machine Intelligence, May 10, 2019 (Co-author)}

  \cvpublication
  {Data-Driven Prediction of Embryo Implantation Probability Using IVF Time-lapse Imaging}
  {MIDL 2020, Jun 1, 2020 (First author)}

  \cvpublication
  {Deep Learning Model for Improving Embryo Selection and Deselection}
  {Human Reproduction (Abstract), Jul 2020 (Co-author)}

  \cvpublication
  {Embryologist Agreement When Assessing Blastocyst Implantation Probability: Is Data-Driven Prediction the Solution to Embryo Assessment Subjectivity?}
  {Human Reproduction, Sep 30, 2022 (Co-author)}

  \cvpublication
  {P-618 Artificial Intelligence (AI) to Optimize Ovarian Stimulation for IVF: From Stimulation Protocols to Total and Mature Oocyte Yield}
  {Human Reproduction (Abstract), Sep 2022 (Co-author)}

  \cvpublication
  {Enhancing Predictive Accuracy in Embryo Implantation: The Bonna Algorithm and its Clinical Implications}
  {Lecture Notes in Computer Science, vol 14976, AIiH 2024, pp 160–171, Aug 15, 2024 (Co-author)}

  \cvpublication
  {Investigating Audience Preferences Within the Hybrid Competitive-Comedic Format of Taskmaster UK}
  {PLOS ONE, 10.1371/journal.pone.0331064, Accepted (Manuscript ID: PONE-D-25-28524R1) (Sole-author)}
\end{cvpublications}

\newpage
\cvsection{Patents}
\begin{cvpatents}
  \cvpatent
    {Auto Range Control for Active Illumination Depth Camera}
    {Intel}
    {US09800795B2, US10451189B2, US10927969B2, US20200072367A1, 2017}

  \cvpatent
    {Code Filters for Coded Light Depth Acquisition in Depth Images}
    {Intel}
    {US09792671B2, US20170178294A1, 2017}

  \cvpatent
    {Morphological and Geometric Edge Filters for Edge Enhancement in Depth Images}
    {Intel}
    {US20170178305A1, 2017}

  \cvpatent
    {Range Reconstruction Using Shape Prior}
    {Intel}
    {US10775501B2, US20180348368A1, 2018–2020}

  \cvpatent
    {Calibration of a Depth Sensing Array Using Color Image Data}
    {Apple Inc}
    {US11914078B2, US20210208262A1, 2021}

  \cvpatent
    {Enhanced Depth Mapping Using Visual Inertial Odometry}
    {Apple Inc}
    {US12153140B2, EP3857259A1, 2021}

  \cvpatent
    {Predicting Embryo Implantation Probability}
    {Embryonics}
    {WO2022009186A1, US20220012873A1, EP4178485A1, 2022}

  \cvpatent
    {Semantically Altering Medical Images}
    {Embryonics}
    {US20220284542A1, 2022}

  \cvpatent
    {Estimating Oocyte Quality}
    {Embryonics}
    {US20220375069A1, 2022}

  \cvpatent
    {Food Processing Systems and Methods}
    {AKA Foods}
    {US20230288392A1, WO2023170639A1, 2023}

  \cvpatent
    {Mixture Modeling Systems and Methods}
    {AKA Foods}
    {US20230289492A1, WO2023170640A1, 2023}

  \cvpatent
    {Molecular Embedding Systems and Methods}
    {AKA Foods}
    {US20230290444A1, WO2023170641A1, 2023}

  \cvpatent
    {Virtual Tasting Systems and Methods}
    {AKA Foods}
    {US20230113613A1, EP4412471A1, 2023}

  \cvpatent
    {Calibrating Texture Cameras Using Features Extracted from Depth Images}
    {Intel}
    {US10540784B2, US20210081781A1, 2020}
\end{cvpatents}

\cvsection{Academic References}
\begin{cvparagraph}
\textbf{Neuroscience \& Research Methodology:}
• Professor Miriam Reiner (Technion) - Research supervisor in Neurocognition and VR Lab; collaborated on brain-computer interface research, EEG analysis, and cognitive neuroscience studies. Close professional and personal acquaintance.
• Professor Hillel Pratt (Technion) - Research collaboration in Evoked Potentials Laboratory; conducted studies on neural signal processing, electrophysiology, and brain response analysis

\textbf{AI \& Geometric Methods:}
• Professor Michael Bronstein (Oxford) - Extensive collaboration on geometric deep learning research; co-researcher in AI methodologies and industry implementation. Close professional and personal acquaintance.
• Professor Alex Bronstein (Vienna) - Long-term research collaborator in geometric computer vision; joint work on AI algorithms. Close professional and personal acquaintance.

\textbf{Computational Biology:}
• Professor Itai Yanai (NYU) - Primary Investigator (PI) for PhD research; co-author on 3 publications including Nature paper; research supervisor in computational evolution and genomics

\textbf{At Reichman University:}
• Professor Doron Friedman (Reichman University) - Met multiple times in academic settings; established professional connection in innovation and media research. Can be used as a general reference of character.
• Dr. Leon Anavy (Reichman University) - Professional collaborator and personal acquaintance; joint research and academic projects

Additional references available upon request.
\end{cvparagraph}

%\input{\sectiondir/skills.tex}
%\input{\sectiondir/interview.tex}

\end{document}
